POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2023-10-14 challenges/linearity.tex

\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \) then
    \( \types{(\Cadd{\Gamma}{l});\Delta}{P} \).
      \item If \( \types{\Gamma; \Delta}{P} \) and \( \tend{\Delta} \) then
    \( \types{\Gamma;(\Cadd{\Delta}{\hastype{x}{\Tend}})}{P} \).
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on the given derivations.
\end{proof}
\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \( \types{\Gamma; (\Delta, x : T)}{P} \) and $x\not\in \freenames{P}$, then \( \types{\Gamma; \Delta}{P} \).
\end{lemma}
\begin{proof}
  By induction on $P$, for each case inverting $\types{\Gamma; (\Delta, x : T)}{P}$ to the uniquely applicable rule.
  \begin{description}
    \item[$P = \Pend$] By inversion on the typing we have $\tend{\Delta}$. Apply T-INACT.
    \item[$P = \Ppar{P_0}{P_1}$] By inversion on the typing we have  $ \Delta, x : T = \Delta_0 , \Delta_1$.
      By cases on which environment $x$ is in, apply IH on that environment. 
     \item[$P = \Presd{y}{z}{P}$] We may assume $x \notin \{y,z\}$. By inversion, $\Delta, x : T = \Delta', y : T_0, z : \dual{T_1}, x : T $. From the assumptions present we can apply IH.\\

       The remaining case are analogous

  \end{description}
\end{proof}

\begin{lemma}[Substitution]\mbox{}
  \label{le:subst}
  If $\types{(\Gamma, l);\Delta}{P}$ and ${\typev{\Gamma}{\hastype{a}{\Tbase}}}$ then
  \( \types{\Gamma;\Delta}{\subst{P}{a}{l}} \).
\end{lemma}
\begin{proof}
  By induction on $P$ inverting  $\types{(\Gamma, l);\Delta}{P}$ in each case.
  \begin{description}
    \item[$P = \Pend$] Immediate after inverting typing.
    \item[$P = \Ppar{P_0}{P_1}$] By inversion on the typing we get derivations for $P_0$ and $P_1$. Apply IH on both.
     \item[$P = \Presd{x}{y}{P}$] Immediate after inverting typing.
     \item[$P = \Pout{x}{v}{P}$] Invert typing. we have  $\typev{(\Gamma,l)}{\hastype{v}{\Tbase}}$.
       We must show $\typev{\Gamma}{\hastype{v}{\Tbase}}$ to build the conclusion with IH and T-OUT. Go by cases on the struture of $\typev{(\Gamma,l)}{\hastype{v}{\Tbase}}$.
     \item[$P = \Pin{x}{v}{P}$] Immediate after inverting typing.

  \end{description}
\end{proof}
\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
  If \( \types{\Gamma;\Delta}{P} \) and \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of a process context involved in the congruence. In the base case where the context is a hole, proceed by case analysis on the \textsl{Sc} rule applied:
  \begin{description}
  \item[Par-Comm/Assoc:] by rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[Par-Inact] right-to-left by
    strengthening~\ref{lemma:strenD}. Vice-versa, by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[Res-Par] Follows Vasco's proof by case analysis on $x : T$ being linear and applying weak and stren accordingly.
  \item[Res-Inact:] use weakening~\ref{lemma:weak} part 2
  \item[Res] order does not matter
  \end{description}
  For the step case, TBD
\end{proof}

\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reduces{P}{Q}$. The cases R-PAR
  and R-RES follows immediately by IH. Case R-STRUCT appeals twice to
  preservation of $\equiv$ (Lemma~\ref{le:presequiv}) and to the
  IH. For R-COM, suppose that T-RES introduces in $\Delta$ the
  assumptions $x : !.U, y : ?\dual U$.  Building the only tree for the
  hypothesis, we know that $\Delta = \Delta_1, \Delta_2, \Delta_3$
  where $\Gamma;\Delta_3 \vdash R$. We also have
  $ \Gamma;\Delta_1 , x : U \vdash P$, ${\cal D}_{2} $ a proof of
  $\Gamma,l;\Delta_2, y : \dual U \vdash Q$ and $\cal V $ is a proof of
  $\Gamma \vdash a : \Tbase$.  From ${\cal D}_{2}$ and $\cal V$ we use the
  substitution Lemma~\ref{le:subst} to obtain $\Gamma;\Delta_2 , y : \dual U \vdash
  Q[a/l]$. We then conclude the proof with rules T-PAR, T-PAR and
  T-RES.
\end{proof}

>>>>

\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \) then
    \( \types{(\Cadd{\Gamma}{l});\Delta}{P} \).
      \item If \( \types{\Gamma; \Delta}{P} \) and \( \tend{\Delta} \) then
    \( \types{\Gamma;(\Cadd{\Delta}{\hastype{x}{\Tend}})}{P} \).
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on the given derivations.
\end{proof}
\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \( \types{\Gamma; (\Delta, x : T)}{P} \) and $x\not\in \freenames{P}$, then \( \types{\Gamma; \Delta}{P} \).
\end{lemma}
\begin{proof}
  By induction on $P$, for each case inverting $\types{\Gamma; (\Delta, x : T)}{P}$ to the uniquely applicable rule.
  \begin{description}
    \item[$P = \Pend$] By inversion on the typing we have $\tend{\Delta}$. Apply \TirName{T-Inact}.
    \item[$P = \Ppar{P_0}{P_1}$] By inversion on the typing we have  $ \Delta, x : T = \Delta_0 , \Delta_1$.
      By cases on which environment $x$ is in, apply IH on that environment. 
    \item[$P = \Presd{y}{z}{P}$] We may assume $x \notin \{y,z\}$. By inversion, $\Delta, x : T = \Delta', y : T_0, z : \dual{T_1}, x : T $. From the assumptions present we can apply IH.
  \end{description}
  The remaining cases are analogous.
\end{proof}

\begin{lemma}[Substitution]\mbox{}
  \label{le:subst}
  If $\types{(\Gamma, l);\Delta}{P}$ and ${\typev{\Gamma}{\hastype{a}{\Tbase}}}$ then
  \( \types{\Gamma;\Delta}{\subst{P}{a}{l}} \).
\end{lemma}
\begin{proof}
  By induction on $P$, inverting  $\types{(\Gamma, l);\Delta}{P}$ in each case.
  \begin{description}
    \item[$P = \Pend$] Immediate after inverting typing.
    \item[$P = \Ppar{P_0}{P_1}$] By inversion on the typing we get derivations for $P_0$ and $P_1$. Apply IH on both.
     \item[$P = \Presd{x}{y}{P}$] Immediate after inverting typing.
     \item[$P = \Pout{x}{v}{P}$] Invert typing. We have $\typev{(\Gamma,l)}{\hastype{v}{\Tbase}}$.
       We must show $\typev{\Gamma}{\hastype{v}{\Tbase}}$ to build the conclusion with IH and \TirName{T-Out}. Go by cases on the structure of $\typev{(\Gamma,l)}{\hastype{v}{\Tbase}}$.
     \item[$P = \Pin{x}{v}{P}$] Immediate after inverting typing.

  \end{description}
\end{proof}
\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
  If \( \types{\Gamma;\Delta}{P} \) and \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of a process context involved in the congruence. In the base case where the context is a hole, proceed by case analysis on the \TirName{Sc} rule applied:
  \begin{description}
  \item[Par-Comm/Assoc:] by rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[Par-Inact] right-to-left by
    strengthening~\ref{lemma:strenD}. Vice-versa, by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[Res-Par] Follows Vasco's proof by case analysis on $x : T$ being linear and applying weak and stren accordingly.
  \item[Res-Inact:] use weakening~\ref{lemma:weak} part 2
  \item[Res] order does not matter
  \end{description}
  For the step case, TBD
\end{proof}

\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reduces{P}{Q}$. The cases \TirName{R-Par} and \TirName{R-Res} follow immediately by IH. Case \TirName{R-Struct} appeals twice to
  preservation of $\equiv$ (Lemma~\ref{le:presequiv}) and to the
  IH. For \TirName{R-Com}, suppose that \TirName{T-Res} introduces in $\Delta$ the
  assumptions $\Cadd{\hastype{x}{\Tout{U}}}{\hastype{y}{\Tin{\dual{U}}}}$.  Building the only tree for the
  hypothesis, we know that $\Delta = \Delta_1, \Delta_2, \Delta_3$
  where $\types{\Gamma;\Delta_3}{R}$. We also have
  $\types{\Gamma;(\Cadd{\Delta_1}{\hastype{x}{U}})}{P}$, ${\cal D}_{2} $ a proof of
  $\types{\Gamma,l;(\Cadd{\Delta_2}{\hastype{y}{\dual{U}}})}{Q}$ and $\cal V $ a proof of
  $\typev{\Gamma}{\hastype{a}{\Tbase}}$.  From ${\cal D}_{2}$ and $\cal V$ we use the
  substitution \cref{le:subst} to obtain $\types{\Gamma;\Cadd{\Delta_2}{\hastype{y}{\dual{U}}}}{\subst{Q}{a}{l}}$. We then conclude the proof with rules \TirName{T-Par} (twice) and \TirName{T-Res}.
\end{proof}
